Aircraft Wildlife collisions

### 2021-07-16 aircraft_wildlife_doc.Rmd
Restructuring and combining plots.

```{r echo=FALSE, fig.asp=1}

@@

@@

#birds seen: remove one wrong entry
which(birds$birds_seen==":-10")
birds$birds_seen[9412] <- NA

factor_seen<-c("2-10", "11-100")
birds$birds_seen <- factor(birds$birds_seen, levels=factor_seen) 
  # drop unused level ":-10" by assigning levels again
levels(birds$birds_seen)

#\newpage
```

@@

```{r echo=FALSE, message=FALSE}
#get rows where birds struck is missing
df <- birds[is.na(birds$birds_struck), 
            c("remarks", "birds_seen", "effect", "birds_struck")]
no_strike <- grepl("NOT A STRIKE|NO STRIKE|NOT A STRKE", df$remarks)
# set birds_struck to 0 for "no strike"
birds[is.na(birds$birds_struck),][no_strike,"birds_struck"] <- "0"

#TODO: get ac mass and num_engs from aircraft model

sum(is.na(birds$atype)) #0
birds[,c("atype", "num_engs", "ac_mass")]
sapply(birds[,c("atype", "num_engs", "ac_mass")], function(x) sum(is.na(x)))

@@

```{r echo=FALSE, fig.asp=1}

above_ground <- birds[as.integer(birds$phase_of_flt)>3 & as.integer(birds$phase_of_flt)<8 & !is.na(birds$phase_of_flt),]
height_by_phase <- ggplot(above_ground) +
  aes(x = phase_of_flt, y = height, fill=phase_of_flt) +
  stat_boxplot(geom="errorbar", width=0.4) +
  geom_boxplot()
# Heights during Descent are on average higher than 
# hypothesis: Collisions happen more often during Descent and are therefore more prominent in the data set?
count_phase <- ggplot(above_ground) +
  aes(x = phase_of_flt, fill=phase_of_flt) +
  geom_bar()

height_by_phase / count_phase
## descent and en-route are about equal

which(birds$height==max(birds$height, na.rm = T))
birds[15053,] # >30000 foot (1 meter = 3,28084 foot)
?birds
```

## Closer look on En Rout and Descent
```{r echo=TRUE, fig.asp=0.5}
#plot height density for Descent and EnRoute
ggplot(birds[birds$phase_of_flt== "Descent" | birds$phase_of_flt=="En Route",]) +
  aes(x = height, group=phase_of_flt,fill=phase_of_flt ) +
  geom_density(adjust=1.5,alpha=.8)

@@

t.test
```{r echo=TRUE, fig.asp=1}
#Two sample t.test for height dependent on phase of flight ("Descent" and "En Route")
t.test(height~phase_of_flt, data = birds[as.integer(birds$phase_of_flt)>4 & as.integer(birds$phase_of_flt)<7 & !is.na(birds$phase_of_flt),])

@@

```{r echo=FALSE, fig.asp=0.5}
#relationship of amount of birds struck with clouds?
birds$effectYes<- as.factor(birds$effect != "None")
birds.filtered<-birds[!is.na(birds$effect) & !is.na(birds$sky),]

prop.table(table(birds.filtered$sky,birds.filtered$birds_struck),1)
ggplot(birds.filtered)+
  aes(x=birds_struck, fill=sky)+
  geom_bar(position = "fill")
```
For 0 birs struck we have highest percentage of no clouds, so clouds might have an affect on avoiding striking birds. Its about 20% higher compared to 1 bird struck.
for birdsstruck by sky we can see that with increasing amount of birds the sky tends to be more more overcast, as no clouds and some clouds decreases

#Clouds with birds seen

@@

```{r echo=FALSE, fig.asp=0.5}
#relationship of amount of birds seen with clouds?
table(birds$birds_seen)

prop.table(table(birds.filtered$sky,birds.filtered$birds_seen),1)
ggplot(birds.filtered[!is.na(birds.filtered$birds_seen),])+
  aes(x=birds_seen, fill=sky)+
  geom_bar(position = "fill")
```

the proportion of some clouds is nearly the same, but pilots do more often see bigger groups of birds (11-100) when its overcast compared to small groups

```{r echo=FALSE, fig.asp=0.5}
#relationship of amount of birds seen with daytime?
table(birds$birds_seen)

prop.table(table(birds.filtered$time_of_day,birds.filtered$birds_seen),1)
ggplot(birds.filtered[!is.na(birds.filtered$birds_seen) & !is.na(birds.filtered$time_of_day),])+
  aes(x=birds_seen, fill=time_of_day)+
  geom_bar(position = "fill")
```
check speed

```{r echo=FALSE, fig.asp=0.5}
# does speed has an influence on seen? the faster the less likely to see ?
spineplot(birds$speed, as.factor(is.na(birds$birds_seen)))
```
The higher the speed, the less likely to see a bird

### Birds struck out of birds seen

```{r echo=FALSE, fig.asp=0.5}
#Create new column
#calculate only when seen is not na
df <- birds[!is.na(birds$birds_seen) & !is.na(birds$birds_struck),]
table(birds$birds_seen, birds$birds_struck)

df$birds_struck <- factor(df$birds_struck)

#mosaic plot better than barplot visualizes also absolute differences
ggplot(df) +
  geom_mosaic(aes(x = product( birds_struck,birds_seen ), fill=birds_struck))
```

```{r echo=FALSE, fig.asp=0.5}
#Create new column
#calculate only when seen is not na
df <- birds[!is.na(birds$birds_seen) & !is.na(birds$birds_struck),]
table(birds$birds_seen, birds$birds_struck)

df$birds_struck <- factor(df$birds_struck)

levels(df$birds_seen)
levels(df$birds_struck)
df$difference_seen_strike <- as.integer(df$birds_struck) < as.integer(df$birds_seen)+2
table(df$difference_seen_strike)

barplot(prop.table(table(df$difference_seen_strike,df$effectYes),2))

ggplot(df) +
  aes(x=difference_seen_strike, fill=effectYes) +
  geom_bar()
```

@@

```{r echo=FALSE, fig.asp=0.5}
ggplot(birds.filtered[!is.na(birds.filtered$birds_struck),]) +
  aes(x = birds_struck, fill = effectYes) +
  geom_bar(position="fill")

#comparison right? because we actually want to compare here the positive impact
ggplot(birds.filtered) +
  aes(x = birds_struck, fill = effect) +
  geom_bar(position="fill")

@@

#mosaic plot better than barplot visualizes also absolute differences
ggplot(birds.filtered[!is.na(birds.filtered$ac_mass),]) +
  geom_mosaic(aes(x = product(ac_mass), fill=effectYes))

# Most collisions during the day #genauer betrachten
ggplot(birds) + 
  aes(x=time_of_day, fill=birds_struck) + 
  geom_bar()

check if clouds have an influence on effect 
hyptohesis: clouds lead to less vision therefore the pilot might not see the birds

```{r echo=FALSE, fig.asp=0.5}
prop.table(table(birds.filtered$sky,birds.filtered$effectYes),1)
ggplot(birds[!is.na(birds$effect) & !is.na(birds$sky),])+
  aes(x=sky, fill=effectYes)+
  geom_bar(position = "fill")
```

results show, that clouds do not have an impact on effect variable, 
the proportion of effect=true is a little higher for some clouds and overcast

check influence of birds seen on effect variable, we assume NA values as if no birds have been seen before the strike occured

```{r echo=FALSE, fig.asp=0.5}
#anteil der effect=true vergleichen when seen and not seen
#...
effectYes_seen <- ggplot(birds[!is.na(birds$effectYes),]) +
  aes(x=is.na(birds_seen), fill=effectYes) +
  geom_bar(position="fill")

#differences in kind of effect?
effect_seen <- ggplot(birds[!is.na(birds$effect) & birds$effect != "None" ,]) +
  aes(x=is.na(birds_seen), fill=effect) +
  geom_bar(position="fill")

effectYes_seen + effect_seen

```
proportion of effect is a little higher when birds were seen

>>>>

```{r echo=FALSE, message= FALSE, fig.asp=1}

@@

factor_sky<-c("No Cloud", "Some Cloud", "Overcast")
birds$sky <- factor(birds$sky, levels=factor_sky)

factor_effect <- c("None","Other","Aborted Take-off","Engine Shut Down","Precautionary Landing")
birds$effect <- factor(birds$effect, levels=factor_effect)

birds$effectYes<- as.factor(birds$effect != "None")

@@

#birds seen: remove one wrong entry
  # which(birds$birds_seen==":-10")
birds$birds_seen[9412] <- NA

factor_seen<-c("None","2-10", "11-100")
birds$birds_seen <- factor(birds$birds_seen, levels=factor_seen) 
  # drop unused level ":-10" by assigning levels again
birds$birds_seenYes <- as.factor(!is.na(birds$birds_seen))
#set Na values in birds seen to "None"
birds[is.na(birds$birds_seen),"birds_seen"] <- "None"
#\newpage
```

@@

```{r echo=FALSE, message=FALSE, fig.asp=0.5}
#get rows where birds struck is missing
df <- birds[is.na(birds$birds_struck), 
            c("remarks", "birds_seen", "effect", "birds_struck")]
no_strike <- grepl("NOT A STRIKE|NO STRIKE|NOT A STRKE", df$remarks)
# set birds_struck to 0 for "no strike"
birds[is.na(birds$birds_struck),][no_strike,"birds_struck"] <- "0"

@@

```{r echo=FALSE, fig.asp=0.5}
birds_strike_phase <- ggplot(birds[!is.na(birds$phase_of_flt) & !is.na(birds$birds_struck),]) +
  aes(x = phase_of_flt, fill=birds_struck) +
  geom_bar()

#relationship of amount of birds struck with clouds?
birds_strike_sky <- ggplot(birds[!is.na(birds$sky) & !is.na(birds$birds_struck),])+
  aes(x=birds_struck, fill=sky)+
  geom_bar(position = "fill")

# Most collisions during the day #genauer betrachten
birds_strike_time <- ggplot(birds[!is.na(birds$time_of_day) & !is.na(birds$birds_struck),]) + 
  aes(x=time_of_day, fill=birds_struck) + 
  geom_bar()

### Birds struck out of birds seen
birds_struck_seen <- ggplot(birds[!is.na(birds$birds_seen) & !is.na(birds$birds_struck),]) +
  geom_mosaic(aes(x = product(birds_seen), fill=birds_struck)) +
  theme(axis.text.y=element_blank(), axis.ticks.y = element_blank())

(birds_strike_phase + birds_strike_sky) / (birds_strike_time + birds_struck_seen)
## descent and en-route are about equal 
```
TODO: add Chi-squared independence test for sky and birds struck
http://www.sthda.com/english/wiki/chi-square-test-of-independence-in-r

the higher the amount of birds struck the more clouds

## Closer look on En Rout and Descent
```{r echo=FALSE, fig.asp=0.5, warning=FALSE}
above_ground <- birds[as.integer(birds$phase_of_flt)>3 & as.integer(birds$phase_of_flt)<8 & !is.na(birds$phase_of_flt),]
height_by_phase <- ggplot(above_ground) +
  aes(x = phase_of_flt, y = height, fill=phase_of_flt) +
  stat_boxplot(geom="errorbar", width=0.4) +
  geom_boxplot()
# Heights during Descent are on average higher than 
# hypothesis: Collisions happen more often during Descent and are therefore more prominent in the data set?

#plot height density for Descent and EnRoute
density <- ggplot(birds[birds$phase_of_flt== "Descent" | birds$phase_of_flt=="En Route",]) +
  aes(x = height, group=phase_of_flt,fill=phase_of_flt ) +
  geom_density(adjust=1.5,alpha=.8)

height_by_phase + density

@@

t.test
```{r echo=FALSE, fig.asp=1}
#Two sample t.test for height dependent on phase of flight ("Descent" and "En Route")
t.test(height~phase_of_flt, data = birds[as.integer(birds$phase_of_flt)>4 & as.integer(birds$phase_of_flt)<7 & !is.na(birds$phase_of_flt),])

```
For 0 birs struck we have highest percentage of no clouds, so clouds might have an affect on avoiding striking birds. Its about 20% higher compared to 1 bird struck.
for birdsstruck by sky we can see that with increasing amount of birds the sky tends to be more more overcast, as no clouds and some clouds decreases

## density birdsstuck by height
```{r echo=FALSE, warning=FALSE, fig.asp=0.5}
#could include only birds from vys categorization
#plot height density for Descent and EnRoute
density <- ggplot(above_ground[!is.na(above_ground$birds_struck),]) +
  aes(x = height, group=birds_struck,fill=birds_struck)+
  geom_density(adjust=1.5,alpha=.6) +
  xlim(0, 3500)
density

@@

## Analysis of birds seen

@@

```{r echo=FALSE, fig.asp=0.5}
#relationship of amount of birds seen with clouds?
birds_seen_plt <- ggplot(birds[!is.na(birds$time_of_day),])+
  aes(x=time_of_day, fill=birds_seen)+
  geom_bar(position = "fill") +
  theme(legend.position="none")

birds_seenNo_sky <- ggplot(birds[!is.na(birds$sky),])+
  aes(x=sky, fill=birds_seen)+
  geom_bar(position="fill")

#relationship of seen or not seen with daytime?
#table(is.na(birds$birds_seen), birds$time_of_day)

birds_seen_plt + birds_seenNo_sky
```
**Time of Day:** Based on the assumption that Missing values in the data for birds_seen stands for the fact that the pilot did not see wildlife before a strike happened, the proportion of birds seen before a strike happened is `r sum(birds$birds_seen!="None")/nrow(birds)`. The contingencytable looks like the following: `addmargins(prop.table(table(birds$birds_seen=="None", birds$time_of_day),2),2,FUN = mean)`.

The proportion of overall birds seen at night is much lower compared to the other times (Dawn, Day, Dusk). And especially large wildlife groups (11-100) are even less likely seen at night compared to smaller groups (2-10).

**Sky:** the proportion of some clouds is nearly the same, but pilots do more often see bigger groups of birds (11-100) when its overcast compared to small groups
contradiction: would have expected to see less birds when amount of clouds increases. but especially the amount of birds in groups from 11-100 increases when the amount of clouds increases.

```{r echo=FALSE, fig.asp=0.5, figures-side, fig.show="hold"}
par(mfrow=c(1,2))
spineplot(birds$height, birds$birds_seen)
spineplot(birds$speed, birds$birds_seen)
```
The higher the speed & height, the less likely to see a bird before collision
for height > 2000 foot probability to see a birds before collision decreases 
for speed > 140 knots probability to see a bird before collision decreases

## effect when birdsstruck < birds seen

```{r echo=FALSE, fig.asp=0.5}
#Create new column
#calculate only when seen is not na
seen_vs_strike <- birds[birds$birds_seen!="None" & !is.na(birds$birds_struck),]
seen_vs_strike$birds_struck <- factor(seen_vs_strike$birds_struck)
seen_vs_strike$birds_seen <- factor(seen_vs_strike$birds_seen)
seen_vs_strike$less_strike_than_seen <- as.factor(as.integer(seen_vs_strike$birds_struck) < as.integer(seen_vs_strike$birds_seen)+2)

#TODO: Why is barplot different from ggplot???
#barplot(prop.table(table(seen_vs_strike$less_strike_than_seen,seen_vs_strike$effectYes),2), xlab = #"less_strike_than_seen")
#prop.table(table(seen_vs_strike$less_strike_than_seen,seen_vs_strike$effectYes),2)

# phase of flight by birdsstruck < birds seen
seen_vs_strike_phase <- ggplot(seen_vs_strike[!is.na(seen_vs_strike$phase_of_flt),]) +
  aes(x=phase_of_flt, fill=less_strike_than_seen) +
  geom_bar(position = "fill")

#ac_mass by leess-strike-than-seen (MOSAIC)
seen_vs_strike_ac_mass <- ggplot(seen_vs_strike[!is.na(seen_vs_strike$ac_mass),]) +
  geom_mosaic(aes(x=product(ac_mass), fill=less_strike_than_seen)) +
   theme(legend.position="none", axis.text.y=element_blank(), axis.ticks.y = element_blank())

#num_engs by leess-strike-than-seen (MOSAIC)
seen_vs_strike_num_engs <- ggplot(seen_vs_strike[!is.na(seen_vs_strike$num_engs),]) +
  geom_mosaic(aes(x=product(num_engs), fill=less_strike_than_seen)) +
   theme(legend.position="none", axis.text.y=element_blank(), axis.ticks.y = element_blank())

seen_vs_strike_phase / (seen_vs_strike_ac_mass + seen_vs_strike_num_engs)
```
the num of engines and ac mass have an influence of how much birds are striked out of the birds seen,
lighter machines and machines with less sngines tend to hit less wildlife out of the seen wildlife. This might be to better maneuvarability of the smaller aircrafts.

@@

```{r echo=FALSE, fig.asp=0.5}
#comparison right? because we actually want to compare here the positive impact
birds_struck_effect <- ggplot(birds[!is.na(birds$birds_struck) & !is.na(birds$effect),]) +
  aes(x = birds_struck, fill = effect) +
  geom_bar(position="fill")

@@

#mosaic plot better than barplot visualizes also absolute differences
birds_effect_acmass <- ggplot(birds[!is.na(birds$ac_mass) & !is.na(birds$effect),]) +
  geom_mosaic(aes(x = product(ac_mass), fill=effect)) +
  theme(legend.position="none", axis.text.y=element_blank(), axis.ticks.y = element_blank())

#birdds_seenYes by effect
birds_effect_seenYes <- ggplot(birds[!is.na(birds$effect),]) +
  geom_mosaic(aes(x = product(birds_seenYes), fill=effect))+
  #theme(legend.title = element_blank())
  theme(legend.position="none", axis.text.y=element_blank(), axis.ticks.y = element_blank())

birds_struck_effect / (birds_effect_acmass + birds_effect_seenYes)

```
check influence of birds seen on effect variable, we assume NA values as if no birds have been seen before the strike occured

proportion of effect is a little higher when birds were seen => contradiction, we would expect to have less amount of effect when pilot is able to see birds before, cause he could
